A general Nullstellensatz for generalised spaces

### 2019-05-02 slides-birmingham2019.tex
Notes finished

  Ingo Blechschmidt \\
  Università di Verona
  \par

  \jnote{1}{\textcolor{red}{This set of slides will be updated soon with more
  annotations.} Right now important references are not properly mentioned, including:
  \begin{itemize}
    \item Marc Bezem, Ulrik Buchholtz and Thierry Coquand. \emph{Syntactic forcing
    models for coherent logic.}
    \item Olivia Caramello. \emph{Universal models and definability.}
    \item Carsten Butz and Peter Johnstone. \emph{Classifying toposes for
    first-order theories.}
  \end{itemize}}
\end{frame}}

@@

    The theory of local rings is the quotient theory of the theory of rings
    obtained by adding the axioms~$(1 = 0 \vdash \bot)$ and~$((\exists z\_
    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1))$.
    (Assuming the Boolean prime ideal theorem, a ring is local in this sense
    iff it local in the usual sense (has exactly one maximal ideal).)

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
\begin{frame}{Exhaustion and extensions}
  \justifying
  \textbf{Theorem.} A first-order formula holds for~$U_\TT$ iff it is
  intuitionistically provable from the axioms of~$\TT$ and
  the scheme
  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
  proves$^{\star\star}$~$\sigma$}. \end{equation}

  \textbf{Theorem.}
  Let~$\TT'$ be a quotient theory
  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
  to~$\Set[\TT']$:
  \begin{quote}
    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
  \end{quote}
  \vspace*{-1.6em}

  \textbf{Theorem.}
  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
  % context~$x\?X$, where any two such formulas are identified if and only
  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
  The morphism~$\mathrm{ev}$ is an isomorphism.
  \[ \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash)
  \longrightarrow P(U_\TT) \]

  \textbf{Theorem.} A higher-order formula holds for~$U_\TT$ iff it is
  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
  the higher-order Nullstellensatz scheme.

  \jnote{1}{
    The first theorem states that the source provided by the Nullstellensatz is
    exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
    the notion of provability$^\star$, which in turn is a strengthening of the
    ordinary notion of provability in geometric logic. We are using it here
    because while the notion of provability$^\star$ can be expressed in the
    internal language of a topos, it cannot be expressed in intuitionistic
    logic.

    The second theorem provides a useful variant of the Nullstellensatz. Its
    assumptions are for instance satisfied if~$\TT$ is the theory of rings
    and~$\TT'$ is the theory of local rings. When applicable, it can be used to
    avoid doubly-internal toposes. It also explains, for instance, why in the
    formulation of synthetic quasicoherence no local rings appear even though
    the relevant topos is the classifying topos of local rings.

    The third and fourth theorems generalise the Nullstellensatz to the
    higher-order setting. The map~$\mathrm{ev}$ takes (the equivalence class
    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.

    Details on all of this are slowly emerging at
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{this
    preprint}.
  }

  \jnote{2}{
    XXX future work
  }
\end{frame}}

>>>>

  Ingo Blechschmidt \\
  Università di Verona
  \par
\end{frame}}

@@

    The theory of local rings is the quotient theory of the theory of rings
    obtained by adding the axioms
    \[ (1 = 0 \vdash \bot) \ \text{and}\ ((\exists z\_
    (x+y)z = 1) \seq{x,y} (\exists z\_ xz = 1) \vee (\exists z\_ yz = 1)). \]
    (Assuming the axiom of choice, a ring is local in this sense
    iff it local in the usual sense (has exactly one maximal ideal).)

@@

{\usebackgroundtemplate{\begin{minipage}{\paperwidth}\vspace*{5.95cm}\hfill\mbox{\includegraphics[width=0.13\paperwidth]{hen}\;\;}\end{minipage}}
\begin{frame}{Exhaustion and extensions}
  \justifying
  \textbf{Theorem 1.} A first-order formula holds for~$U_\TT$ iff it is
  intuitionistically provable from the axioms of~$\TT$ and
  the scheme
  \begin{equation}\tag{Nullstellensatz} \speak{$\sigma$ holds} \implies \speak{$\ul{\TT}/U_\TT$
  proves$^{\star\star}$~$\sigma$}. \end{equation}

  \textbf{Theorem 2.}
  Let~$\TT'$ be a quotient theory
  of~$\TT$. Assume that~$U_\TT$ is contained in the subtopos~$\Set[\TT']$. Then internally
  to~$\Set[\TT']$:

  \vspace*{-0.4em}
  \begin{quote}
    \emph{A geometric$^\star$ sequent~$\sigma$ with Horn consequent holds for~$U_{\TT'}$
    iff~$\ul{\TT}/U_\TT$ proves$^\star$~$\sigma$.}
  \end{quote}

  \textbf{Theorem 3.}
  % Let~$\mathrm{Form}^\star_{x\?X}(\ul{\TT}/U_\TT)/(\dashv\vdash_{x\?X})$ be the~$\Set[\TT]$-object of
  % geometric$^\star$ formulas over the signature of~$\ul{\TT}/U_\TT$ in the
  % context~$x\?X$, where any two such formulas are identified if and only
  % if~$\ul{\TT}/U_\TT$ proves them equivalent.
  The morphism~$\mathrm{ev}$ below is an isomorphism.
  \begin{equation}\label{eq:hon}\tag{$\star$}
    \mathrm{ev} : \mathrm{FunctFormulas}^\star(\ul{\TT}/U_\TT)/(\dashv\vdash) \longrightarrow P(U_\TT)
  \end{equation}

  \textbf{Theorem 4.} A higher-order formula holds for~$U_\TT$ iff it is
  provable in intuitionistic higher-order logic from the axioms of~$\TT$ and
  the higher-order Nullstellensatz scheme~($\star$).

  \jnote{1}{
    Theorem~1 states that the source provided by the Nullstellensatz is
    exhaustive. The notion of provability$^{\star\star}$ is a strengthening of
    the notion of provability$^\star$, which in turn is a strengthening of the
    ordinary notion of provability in geometric logic. We are using it here
    because while the notion of provability$^\star$ can be expressed in the
    internal language of a topos, it cannot be expressed in intuitionistic
    logic.

    Theorem~2 provides a useful variant of the Nullstellensatz. Its
    assumptions are for instance satisfied if~$\TT$ is the theory of rings
    and~$\TT'$ is the theory of local rings. When applicable, it can be used to
    avoid doubly-internal toposes. It also explains, for instance, why in the
    formulation of synthetic quasicoherence no local rings appear even though
    the relevant topos is the classifying topos of local rings.

    Theorems~3 and~4 generalise the Nullstellensatz to the
    higher-order setting. The map~$\mathrm{ev}$ maps (the equivalence class
    of) a~$\ul{\TT}/U_\TT$-provably$^\star$ geometric$^\star$ formula~$\theta$
    in one free variable to the subset~$\{ x \? U_\TT \,|\, \theta(x) \}$.

    Written details on all of this are
    \fixedhref{https://rawgit.com/iblech/internal-methods/master/paper-qcoh.pdf}{slowly
    emerging}.
  }

  \jnote{2}{
    The Nullstellensatz is related to several precursors. A corollary of the
    Nullstellensatz is that, over the first-order theory validated by~$U_\TT$, any
    first-order formula is in fact logically equivalent to a geometric formula.
    This corollary has already been observed by Carsten Butz and Peter
    Johnstone in their paper
    \fixedhref{http://citeseerx.ist.psu.edu/viewdoc/download?doi=10.1.1.16.4102&rep=rep1&type=pdf}{Classifying
    toposes for first-order theories} (Lemma~4.2 there). At that point, a
    characterisation of the first-order formulas in the general case, of the
    form as in Theorem~1, was still missing.

    Theorem~3 is a relativisation of Olivia Caramello's completeness theorem,
    Theorem~2.4(ii) in her paper
    \fixedhref{https://arxiv.org/abs/0906.3061}{Universal models and
    definability}. The passage from the external to the internal phrasing
    requires going from~$\TT$ to~$\ul{\TT}/U_\TT$.
  }

  \jnote{2}{
    Plans for the future include:
    \begin{itemize}\justifying
      \item Developing an Agda library for dealing with the internal language
      of toposes and related kinds of categories, employing Agda's
      meta-programming facilities; with such a library at hand,
      formalising the Nullstellensatz in Agda.
      \item Exploring Nullstellensatz-style results for arithmetic universes.
      \item Applying the Nullstellensatz in constructive algebra and algebraic
      geometry, along the lines of generic freeness and synthetic algebraic
      geometry. Most of the toposes in geometric use are actually uncharted
      territory from a logical point of view.
    \end{itemize}
  }
\end{frame}}
